---
title: 'Introduction to R'
subtitle: 'Part 1 of Discussion Section SA4 in BIOL119-2025'
author: 'Author: Your Name'
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: yes
    toc_depth: 3
    fig_caption: yes
    code_folding: show
    number_sections: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
fontsize: 14pt
bibliography: bibtex.bib
type: docs
weight: 304
editor_options: 
  chunk_output_type: console
---

<!---
- Compile from command-line
Rscript -e "rmarkdown::render('Intro2R.Rmd', c('html_document'), clean=FALSE)"
-->

# Get started 

The following provides a brief instruction how to access Posit/Rstudio Cloud, clone a
pre-generated project for this exercise, and run the provided analysis code. This week'ss
computer section will first introduce the basics of R and then use it for phylogenetics 
analysis.

## Initialize project 

The following will initialize the assignment project required for this tutorial. 

1. When using Posit Cloud the first time, create an account [here](https://posit.cloud/). The free account is sufficient. If you have an account log in to Posit Cloud on the same page.
2. Next click the <span style="background-color: #f1c40f;">Assignment Link</span></a> provided under the same name
   of the corresponding _Canvas Assignment_ page
   [here](https://elearn.ucr.edu/courses/192776/assignments/925649). Note
   the latter is private to this class.
3. Start the pre-generated assignment by clicking `BIOL119-2025-SA4`. This will create a copy of the assignment in a student's account.  

## Load R code and run it

Once you R session in Posit/RStudio Cloud has opened, the R code along with the human readable text in this document can be accessed and executed as follows:

1. Open the `Rmd` source file for this document by selecting in the top menu: `File` ->
   `Open File` -> `Intro2R.Rmd`. Alternatively, one can click the name of this 
   file in Posit/RStudio Cloud's file browser, usually located in the window on the bottom right. Either option
   will open the `Rmd` in the code editor window usually located in the top left window.
2. Execute code of the `Rmd` file line by line by placing the cursor on a line of R code 
   and then pressing on the keyboard `Ctrl + Enter`. This will send the corresponding code line
   to the R console (located on bottom left) and execute (run) the code. It is important to execute the
   lines in the order given in the `Rmd` file. Note, `Rmd` files are special `R` files that contain 
   both human readable text in markdown format as well as R code located in so called code chunk boxes 
   that are initialized and terminated with a tag line starting with three backtick (\`) characters. If an 
   `R` script instead of an `Rmd` script is used then everything works the same. However, 
   in an `R` script human readable text has to be initialized on each line with a comment (`#`) sign.

Alternatively, one can execute all code in this document at once simply by
pressing the triangle next to the Knit button above the code editor window in
Posit/RStudio, and then selecting `Knit to HTML`. This will not only execute the
code, but also regenerate this HTML document and update all output, tables and
figures generated by the evaluated R code chunks accordingly.  Similarly, one
can generate a PDF document instead. The environment that makes this possible
is called [R Markdown](https://rmarkdown.rstudio.com/).

# Import Libraries
R packages required for this exercise will be loaded with the `library` command. 

```{r example1, eval=TRUE, message=FALSE}
suppressPackageStartupMessages({
    library(tidyr)
    library(ggplot2)
})
```

# Import of Table 

The following provides a very brief introduction to R using a tabular file as an example [@manualr]. 
As tabular file, a test data set within R called `iris` will be written to a file. After 
executing the following `write.table` command, a new file named `iris.tab` will appear in 
the File browser panel. From within RStudio Cloud this file can be downloaded with the 
Export option in the same panel. For learning purposes, also try this example with your
own tabular file, e.g. exported from Excel or Google Sheets.

```{r write_iris, eval=TRUE, message=FALSE}
write.table(iris, file="iris.tab", row.names=FALSE, quote=FALSE, sep="\t") 
```

Next, the exported sample file will be imported into R with the `read.delim` function. 
The data will be assigned to an object named `df`. The assignment operator `<-` will be
used for this. The class of the resulting object is a `data.frame`. Its dimensions can
be checked with the `dim` function, returning here 150 rows and 5 columns. To print only
the first 4 rows to the screen, one can use the subsetting operator `[1:4 , ]`.
With `[1:4, 1:2]` one can subset the object to the first 4 rows and first 2
columns.

```{r import_and_inspect, eval=TRUE, message=FALSE}
df <- read.delim("iris.tab")
class(df)
dim(df)
df[1:4,]
df[1:4, 1:2]
```

## Nicer tables

To print nicer tables, in the R Markdown `html` report, the `kable` function from the knitr package can be used.

```{r}
library(knitr)
kable(df[1:4,])
```


# Basic Calculations

The following performs some basic calculations on the imported tabular file. First, the `df`
is converted to a numeric `matrix` named `ma`, and then the mean (sum) is calculated for the 
first row as well as for all rows using the `mean` and `rowMeans` functions, respectively.


```{r calculations, eval=TRUE, message=FALSE}
ma <- data.matrix(df[,-5])
class(ma)
ma[1:4,]
mean(ma[1,])
rowMeans(ma)[1:10] # prints first 10 of 150 values
rowSums(ma)[1:10] # prints first 10 of 150 values
```

# Basic Plots

The following creates a simple barplot for the first row in `ma` using the `barplot` function of base R.  

```{r plotting, eval=TRUE, message=FALSE}
barplot(ma[1,])
```

# Nicer Plots with `ggplot2`

Much nicer plots can be generated with the `ggplot2` package. Additional examples for `ggplot2` functionalities can be found [here](https://rb.gy/2f9pr).

```{r ggplot_plotting, eval=TRUE, message=FALSE}
library(ggplot2)
df_mean <- aggregate(df[,1:4], by=list(Species=df$Species), FUN=mean)
df_mean
df_mean <- tidyr::pivot_longer(df_mean, !Species, names_to="Samples", values_to="Values")
p <- ggplot(df_mean, aes(Samples, Values, fill = Species)) + 
        geom_bar(position="dodge", stat="identity")
p
```

Plots can be saved to files in different graphics formats using the R functions with the corresponding
names, including: `pdf`, `svg`, `tiff`, `png`, `jpg`, etc. The file name is specified under the
the first argument (here `myplot.png`).

```{r save_plot, eval=TRUE, message=FALSE}
png("myplot.png")
p
dev.off()
```

# Important Utilities

## Help

The help files for R functions can be opened by typing `?` followed by the name of the functions.
For instance, `?t.test` will open the help file for the Student's t-Test function in R.

## Packages

Many R packages (libraries) can be installed with the `install.packages` or the `BiocManager::install` functions. Before a package can be used, it needs to be loaded with the `library` function. Most of the relevant details for package installs are explained [here](https://rb.gy/8yl3y).

## Tutorials

Many tutorials and books are available for learning R. Here is a detailed [tutorial](https://girke.bioinformatics.ucr.edu/GEN242/tutorials/rbasics/rbasics/) 
of the UCR GEN242 class. Additional tutorials are linked on this  [page](https://pairach.com/2012/02/26/r-tutorials-from-universities-around-the-world/).


# Session Info

```{r sessionInfo}
sessionInfo()
```

# References
